---
title: "Download Curated Metagenomic Data"
output: html_notebook
---

Jacobo de la Cuesta-Zuluaga. December 2019.

The aim of this notebook is to obtain the latest samples from the `curatedMetagenomicData` package, filter it and prepare tables for download using the LLMGQC pipeline.

*Note* that this notebook is using the development version of `curatedMetagenomicData` and **requires the development version of `R`, i.e. `v.4.0`**

# Lib
```{r message=FALSE}
library(curatedMetagenomicData)
library(tidyverse)
```

# Init
```{r}
# Work dir
work_dir = "/ebio/abt3_projects/small_projects/jdelacuesta/public_data_retrieval/CurMetDat-metagenomes"
```

# Curated metagenome data
The complete metadata table is stored in a df called `combined_metadata`.

## Explore full metadata
```{r}
# Head
combined_metadata %>% head
```

```{r}
# Studies and number of samples per study
combined_metadata %>% count(dataset_name)
```

# Export files
```{r}
dir.create(file.path(work_dir, "files"))
# Complete CurMetDat data frame
write_tsv(combined_metadata, file.path(work_dir, "files", "Complete_CurMetDat.tsv")) 
```


